---
title: "Regression Analysis"
author: "Alexis Alberto Zúñiga Alonso"
date: today
format:
  html:
    theme: cosmo
    toc: true
    code-fold: show
    code-tools: true
    embed-resources: true
    fig-width: 8
execute: 
  warning: true
  message: true
  echo: true
---

## 1. Introduction

### 1.1. Project Libraries and Installation
```{r}
# Project libraries
libs <- c("genridge", "tidyverse", "skimr", "corrplot", "car",
          "lmtest", "glmnet", "MASS", "performance", "see", "gt", "psych")

# Install only the missing ones
missing <- setdiff(libs, rownames(installed.packages()))
if (length(missing) > 0) {
  install.packages(missing, repos = "https://cloud.r-project.org")
}

# Load all of them
invisible(lapply(libs, library, character.only = TRUE))
```

### 1.2. Dataset Analysis

#### 1.2.1. Importing the dataset
The dataset to be used in this activity will be `prostate` from the `genridge` library. This dataset contains the following structure:

```{r}
#| eval: false
# Information about the library
library(genridge)
?prostate
```

#### 1.2.2. Dataset
```{r}
# Dataframe creation
data(prostate, package="genridge")
df <- prostate |> dplyr::select(-train)
# Stringify dataset
str(df)
# Dataset summary
print(summary(df))
# Skim through the data
skimr::skim(df)
# Head of the dataset
print(head(df))
# Tail of the dataset
print(tail(df))
# Shape / Dimensions of the Dataset
print(dim(df))
```

### 1.3. Descriptive Statistics
```{r}
# Descriptive statistics with psych
psych::describe(df) |>
  round(2) |>
  gt::gt(rownames_to_stub = TRUE)
```

#### 1.3.1. Histograms

```{r}
hist(df$lcavol,
     main = "Distribution of the Cancer Volume",
     xlab = "log(cancer volume)",
     col = "coral",
     border = "red",
)
hist(df$lweight,
     main = "Distribution of the Prostate Weight",
     xlab = "log(prostate weight)",
     col = "coral",
     border = "red",
)
hist(df$age,
     main = "Distribution of Ages",
     xlab = "ages",
     col = "coral",
     border = "red",
)
hist(df$lbph,
     main = "Distribution of Benign Prostatic Hyperplasia Amount",
     xlab = "log(benign prostatic hyperplasia amount)",
     col = "coral",
     border = "red",
)
hist(df$svi,
     main = "Distribution of the Seminal Vesicle Invasion",
     xlab = "seminal vesicle invasion",
     col = "coral",
     border = "red",
)
hist(df$lcp,
     main = "Distribution of the Capsular Penetration",
     xlab = "log(capsular penetration)",
     col = "coral",
     border = "red",
)
hist(df$gleason,
     main = "Distribution of Gleason Scores",
     xlab = "Gleason scores",
     col = "coral",
     border = "red",
)
hist(df$pgg45,
     main = "Distribution of Percentage Gleason Scores 4 or 5",
     xlab = "percentage Gleason scores 4 or 5",
     col = "coral",
     border = "red",
)
hist(df$lpsa,
     main = "Distribution of Prostate Specific Antigen",
     xlab = "log(prostate specific antigen)",
     col = "coral",
     border = "red",
)
```

#### 1.3.2 Normality Test
Shapiro-Wilk test for normality on the variables in the dataset.
```{r}
# Shapiro-Wilk on lcavol
shapiro.test(df$lcavol)
# Shapiro-Wilk on lweight
shapiro.test(df$lweight)
# Shapiro-Wilk on age
shapiro.test(df$age)
# Shapiro-Wilk on lbph
shapiro.test(df$lbph)
# Shapiro-Wilk on svi
shapiro.test(df$svi)
# Shapiro-Wilk on lcp
shapiro.test(df$lcp)
# Shapiro-Wilk on gleason
shapiro.test(df$gleason)
# Shapiro-Wilk on pgg45
shapiro.test(df$pgg45)
# Shapiro-Wilk on lpsa
shapiro.test(df$lpsa)
```

**Observations:**

- The variables that follow a **normal distribution** according to the results given by the Shapiro-Wilk test are: `lcavol`, `lweight`, and `lpsa`. The rest of variables **do not** follow a normal distribution.

### 1.4. Correlation Matrix

```{r}
corrplot(cor(df), method = "color", addCoef.col = "black")
```
