---
title: "Regression Analysis"
author: "Alexis Alberto Zúñiga Alonso"
date: today
format:
  html:
    theme: cosmo
    toc: true
    code-fold: show
    code-tools: true
    embed-resources: true
    fig-width: 8
  typst:
    toc: true
    fig-width: 7
execute:
  warning: true
  message: true
  echo: true
---

## 1. Introduction

### 1.1. Project Libraries and Installation
```{r}
#| message: false
# Project libraries
libs <- c("genridge", "tidyverse", "skimr", "corrplot", "car",
          "lmtest", "glmnet", "MASS", "performance", "see", "gt", "psych")

# Install only the missing ones
missing <- setdiff(libs, rownames(installed.packages()))
if (length(missing) > 0) {
  install.packages(missing, repos = "https://cloud.r-project.org")
}

# Load all of them
invisible(lapply(libs, library, character.only = TRUE))
```

### 1.2. Dataset Analysis

#### 1.2.1. Importing the dataset
The dataset to be used in this activity will be `prostate` from the `genridge` library. This dataset contains the following structure:

```{r}
#| eval: false
# Information about the library
library(genridge)
?prostate
```

#### 1.2.2. Dataset
```{r}
# Dataframe creation
data(prostate, package="genridge")
df <- prostate |> dplyr::select(-train)
# Stringify dataset
str(df)
# Dataset summary
print(summary(df))
# Skim through the data
skimr::skim(df)
# Head of the dataset
print(head(df))
# Tail of the dataset
print(tail(df))
# Shape / Dimensions of the Dataset
print(dim(df))
```

### 1.3. Descriptive Statistics
```{r}
# Descriptive statistics with psych
psych::describe(df) |>
  as.data.frame() |>
  dplyr::select(n, mean, sd, median, min, max, range, skew, kurtosis, se) |>
  round(2) |>
  gt::gt(rownames_to_stub = TRUE)
```

#### 1.3.1. Histograms

```{r}
df |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 20, fill = "coral", color = "red") +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Distribution of every variable", x = NULL, y = "Frequency") +
  theme_minimal()
```

#### 1.3.2 Normality Test
Shapiro-Wilk test for normality on the variables in the dataset.
```{r}
# Shapiro-Wilk on lcavol
shapiro.test(df$lcavol)
# Shapiro-Wilk on lweight
shapiro.test(df$lweight)
# Shapiro-Wilk on age
shapiro.test(df$age)
# Shapiro-Wilk on lbph
shapiro.test(df$lbph)
# Shapiro-Wilk on svi
shapiro.test(df$svi)
# Shapiro-Wilk on lcp
shapiro.test(df$lcp)
# Shapiro-Wilk on gleason
shapiro.test(df$gleason)
# Shapiro-Wilk on pgg45
shapiro.test(df$pgg45)
# Shapiro-Wilk on lpsa
shapiro.test(df$lpsa)
```

**Observations:**

- The variables that follow a **normal distribution** according to the results given by the Shapiro-Wilk test are: `lcavol`, `lweight`, and `lpsa`. The rest of variables **do not** follow a normal distribution.

### 1.4. Correlation Matrix

```{r}
#| fig-height: 7
corrplot(cor(df), method = "color", addCoef.col = "black",
         number.cex = 0.75, tl.col = "black")
```

## 2. Train / Test Split

The seed is fixed with `set.seed(123)` so the split is reproducible. 70% of the observations are used to train the models and the remaining 30% to evaluate their performance.

```{r}
set.seed(123)
idx   <- sample(nrow(df), 0.7 * nrow(df))
train <- df[idx, ]
test  <- df[-idx, ]

cat("Train:", nrow(train), "observations\n")
cat("Test: ", nrow(test), "observations\n")
```

## 3. Multiple Linear Regression (Full Model)

`lpsa` is the response variable and all the other variables are predictors.

```{r}
full_model <- lm(lpsa ~ ., data = train)
summary(full_model)
```

### 3.1. Multicollinearity (VIF)

```{r}
vif(full_model)
```

**Observations:**

- Only `lcavol`, `lweight` and `svi` are significant at the 5% level; the other predictors do not contribute significantly once the rest are in the model.
- The highest VIFs correspond to `lcp` and `pgg45`. No VIF is above 5, so multicollinearity is moderate, but the correlated predictors (`lcavol`–`lcp`, `gleason`–`pgg45`) justify applying variable selection methods.

## 4. Variable Selection Methods

### 4.1. Forward, Backward and Stepwise

All three methods use the AIC as the selection criterion. Forward and stepwise start from the null model (intercept only); backward starts from the full model.

```{r}
null_model <- lm(lpsa ~ 1, data = train)

forward_model  <- step(null_model, scope = formula(full_model),
                       direction = "forward", trace = 0)
backward_model <- step(full_model, direction = "backward", trace = 0)
stepwise_model <- step(null_model, scope = formula(full_model),
                       direction = "both", trace = 0)

formula(forward_model)
formula(backward_model)
formula(stepwise_model)
```

### 4.2. Lasso, Ridge and Elastic Net

`glmnet` requires a numeric matrix of predictors. The penalty parameter $\lambda$ is chosen with 10-fold cross-validation (`cv.glmnet`), using the value that minimizes the cross-validation error (`lambda.min`). The seed is set before each call so the folds are reproducible.

- **Lasso:** $\alpha = 1$
- **Ridge:** $\alpha = 0$
- **Elastic Net:** $\alpha = 0.5$

```{r}
x_train <- model.matrix(lpsa ~ ., data = train)[, -1]
y_train <- train$lpsa
x_test  <- model.matrix(lpsa ~ ., data = test)[, -1]

set.seed(123)
lasso_cv <- cv.glmnet(x_train, y_train, alpha = 1)
set.seed(123)
ridge_cv <- cv.glmnet(x_train, y_train, alpha = 0)
set.seed(123)
enet_cv  <- cv.glmnet(x_train, y_train, alpha = 0.5)

par(mfrow = c(1, 3))
plot(lasso_cv, main = "Lasso")
plot(ridge_cv, main = "Ridge")
plot(enet_cv,  main = "Elastic Net")
par(mfrow = c(1, 1))
```

Coefficients of each penalized model (a coefficient of 0 means the variable was eliminated):

```{r}
penalized_coefs <- cbind(
  as.matrix(coef(lasso_cv, s = "lambda.min")),
  as.matrix(coef(ridge_cv, s = "lambda.min")),
  as.matrix(coef(enet_cv,  s = "lambda.min"))
)
colnames(penalized_coefs) <- c("Lasso", "Ridge", "Elastic Net")
round(penalized_coefs, 4)
```

## 5. Model Comparison (RMSE and MAE)

All models are evaluated on the **test set**, which was not used during training.

```{r}
rmse <- function(y, pred) sqrt(mean((y - pred)^2))
mae  <- function(y, pred) mean(abs(y - pred))

predictions <- list(
  "Full model"  = predict(full_model, test),
  "Forward"     = predict(forward_model, test),
  "Backward"    = predict(backward_model, test),
  "Stepwise"    = predict(stepwise_model, test),
  "Lasso"       = as.numeric(predict(lasso_cv, x_test, s = "lambda.min")),
  "Ridge"       = as.numeric(predict(ridge_cv, x_test, s = "lambda.min")),
  "Elastic Net" = as.numeric(predict(enet_cv,  x_test, s = "lambda.min"))
)

metrics <- tibble(
  Model = names(predictions),
  RMSE  = map_dbl(predictions, function(p) rmse(test$lpsa, p)),
  MAE   = map_dbl(predictions, function(p) mae(test$lpsa, p))
)

metrics |>
  arrange(RMSE) |>
  gt() |>
  fmt_number(columns = c(RMSE, MAE), decimals = 4)
```

**Observations:**

- Among the variable selection methods, **Backward** has the lowest RMSE, and its MAE is practically tied with Lasso's (difference smaller than 0.003). It is therefore the method with the best overall performance, and it is the model analyzed in the next section.
- The full model obtains slightly lower errors than all the selection methods. With only 30 test observations these differences are small; Backward achieves similar performance with half of the predictors (`lcavol`, `lweight`, `svi`, `pgg45`), which makes it simpler to interpret.

## 6. Analysis of the Best Model (Backward)

```{r}
best_model <- backward_model
summary(best_model)
```

### 6.1. Significance of the Model (ANOVA and F Statistic)

- $H_0$: all the coefficients of the predictors are equal to zero.
- $H_1$: at least one coefficient is different from zero.

```{r}
anova(best_model)

f <- summary(best_model)$fstatistic
p_value <- pf(f["value"], f["numdf"], f["dendf"], lower.tail = FALSE)
cat("F =", round(f["value"], 3), "on", f["numdf"], "and", f["dendf"],
    "DF,  p-value =", signif(p_value, 3), "\n")
```

**Observations:** the p-value of the F statistic is far below 0.05, so $H_0$ is rejected: the model is globally significant. In the ANOVA table, `lcavol`, `lweight` and `svi` are significant, while `pgg45` is not significant at the 5% level, although the AIC kept it in the model.

### 6.2. Assumption Checks

```{r}
#| fig-height: 8
par(mfrow = c(2, 2))
plot(best_model)
par(mfrow = c(1, 1))
```

#### 6.2.1. Normality of the Residuals

- $H_0$: the residuals follow a normal distribution.
- $H_1$: the residuals do not follow a normal distribution.

```{r}
qqnorm(residuals(best_model), main = "Normal Q-Q Plot of the Residuals")
qqline(residuals(best_model), col = "red")
shapiro.test(residuals(best_model))
```

**Observations:** the points of the Q-Q plot follow the reference line and the p-value of the Shapiro-Wilk test is greater than 0.05, so $H_0$ is not rejected: there is no evidence against the normality of the residuals.

#### 6.2.2. Homoscedasticity

- $H_0$: the variance of the residuals is constant (homoscedasticity).
- $H_1$: the variance of the residuals is not constant (heteroscedasticity).

```{r}
plot(fitted(best_model), residuals(best_model),
     xlab = "Fitted values", ylab = "Residuals",
     main = "Residuals vs Fitted", pch = 19, col = "coral")
abline(h = 0, col = "red", lty = 2)
bptest(best_model)
```

**Observations:** the residuals are randomly scattered around zero without a funnel shape, and the p-value of the Breusch-Pagan test is greater than 0.05, so $H_0$ is not rejected: the variance of the residuals is constant.

#### 6.2.3. Independence

- $H_0$: the residuals are not autocorrelated (independence).
- $H_1$: the residuals are autocorrelated.

```{r}
plot(residuals(best_model), type = "b",
     xlab = "Observation order", ylab = "Residuals",
     main = "Residuals vs Order", pch = 19, col = "coral")
abline(h = 0, col = "red", lty = 2)
dwtest(best_model)
```

**Observations:** the residuals show no pattern with respect to the observation order, and the Durbin-Watson statistic is close to 2 with a p-value greater than 0.05, so $H_0$ is not rejected: the residuals are independent.

## 7. Conclusions

- The Backward model, with `lcavol`, `lweight`, `svi` and `pgg45`, had the best overall performance among the variable selection methods on the test set.
- The model is globally significant according to the ANOVA F test.
- The assumptions of normality, homoscedasticity and independence of the residuals are satisfied, so the inference of the model is valid.
- The main predictor of `lpsa` is `lcavol` (the log of the cancer volume), followed by `lweight` and `svi`.
